Categorical Foundations of E-Graphs With Binding

### 2025-07-11 POPL2026/paper/sections/introduction.tex
cosmetic changes

String diagrams provide a topological framework for understanding monoidal categories, representing objects as \emph{strings} and morphisms as \emph{nodes}, where input strings flow in and output strings flow out.
In this article, our string diagrams are oriented left-to-right, bottom-to-top to make the pictures look more similar to those commonly drawn for e-graphs.
The structure of symmetric monoidal categories is absorbed into the graphical syntax, with the composition and tensoring represented by juxtaposition, and symmetries as weaving of strings.
This visual approach makes hides the bureacracy in complex operations and reveals insights sometimes obscured by traditional one-dimensional symbolic notation.
Originating from notations by~\citet{hotzsd} (for automata) and~\citet{penrose1984spinors} (for tensor networks), extensions of string diagrammatic calculi exist for various forms of monoidal categories and even bicategories~\cite{Selinger_2010}, and provide a mathematically rigorous reasoning toolkit~\cite{joyal_geometry_1991}.

@@

\citet{tiurin2025equivalencehypergraphsdporewriting} shows that e-graphs over signature $\Sigma$ can be represented as morphisms of a free cartesian symmetric monoidal semilattice-enriched category over the same signature.
We argue that restricting the categorical domain to free \textbf{closed} symmetric monoidal semilattice-enriched categories gives rise to structures that naturally support binding and equivalence classes of morphisms.
That is, morphisms (string diagrams) of such category can both encode variable binding (using the closed structure) and equivalence (using the semilattice enrichment) between subterms (sub-diagrams).
We call these string diagrams, or more precisely, the combinatorial representation of them, \emph{closed e-hypergraphs}.
First, we define closed enriched $\Sigma$-terms which generate a theory with equivalences and bindings.
Then we define what e-hypergraphs are and provide the interpretation of closed enriched $\Sigma$-terms as cospans of e-hypergraphs (and by doing this we also formalise the corresponding string diagrams as concrete combinatorial objects).
Next we formulate the DPO rewriting for e-hypergraphs and show that this notion of rewriting is sound and complete with respect to rewriting of closed enriched $\Sigma$-terms modulo SMC equations.
We finish the paper with a description of what benefits the e-hypergraph representation may bring for equality saturation in the context of $\lambda$-calculus with explicit substitution compared to conventional (slotted) e-graphs.

>>>>

String diagrams provide a topological framework for understanding monoidal categories, representing objects as \emph{strings} and morphisms as \emph{nodes}, where input strings flow in and output strings flow out.
In this article, our string diagrams are oriented left-to-right, bottom-to-top to make the pictures look more similar to those commonly drawn for e-graphs.
The structure of symmetric monoidal categories is absorbed into the graphical syntax, with the composition and tensoring represented by juxtaposition, and symmetries as weaving of strings.
This visual approach hides the bureacracy in complex operations and reveals insights sometimes obscured by traditional one-dimensional symbolic notation.
Originating from notations by~\citet{hotzsd} (for automata) and~\citet{penrose1984spinors} (for tensor networks), extensions of string diagrammatic calculi exist for various forms of monoidal categories and even bicategories~\cite{Selinger_2010}, and provide a mathematically rigorous reasoning toolkit~\cite{joyal_geometry_1991}.

@@

\citet{tiurin2025equivalencehypergraphsdporewriting} shows that e-graphs over signature $\Sigma$ can be represented as morphisms of a free Cartesian symmetric monoidal semilattice-enriched category over the same signature.
We argue that restricting the categorical domain to free \textbf{closed} symmetric monoidal semilattice-enriched categories gives rise to structures that naturally support binding and equivalence classes of morphisms.
That is, morphisms (string diagrams) of such category can both encode variable binding (using the closed structure) and equivalence (using the semilattice enrichment) between subterms (sub-diagrams).
We call these string diagrams, or more precisely, the combinatorial representation of them, \emph{closed e-hypergraphs}.
First, we define closed enriched $\Sigma$-terms which generate a theory with equivalences and bindings.
Then we define what e-hypergraphs are and provide the interpretation of closed enriched $\Sigma$-terms as cospans of e-hypergraphs (and by doing this we also formalise the corresponding string diagrams as concrete combinatorial objects).
Next we formulate the DPO rewriting for e-hypergraphs and show that this notion of rewriting is sound and complete with respect to rewriting of closed enriched $\Sigma$-terms modulo SMC equations.
We finish the paper with a description of what benefits the e-hypergraph representation may bring for equality saturation in the context of $\lambda$-calculus with explicit substitution compared to conventional (slotted) e-graphs.

### 2025-07-11 POPL2026/paper/sections/lambda-calculus.tex
cosmetic changes

Note that, in contrast to conventional presentations of the $\lambda$-calculus as a language of (slotted~\cite[Example 1]{slotted-egraphs}) e-graphs~\cite[Figure 10]{EggPaper}, here there is no encoding of the term formers for application, abstraction, and explicit substitution.
Instead, these constructs are handled ambiently by our semantic setting: any closed symmetric monoidal category (and in particular, $\catname{CS}(\Sigma, \mathcal{E})$ for any set of equations $\mathcal{E}$) admits an internal language in the multiplicative fragment of intuitionistic linear logic (a particular type of linear $\lambda$-calculus)~\cite[§ 1.7.3]{abramskyIntroductionCategoriesCategorical2010}.
In other words, with respect to~\autoref{fig:egraph-strings}, this allows us to \emph{natively} represent abstraction by the rounded boxes of ($\Lambda(f)$), and application by the $\textsf{ev}$ operation with $\beta$- and $\eta$-conversion arising via equality of morphisms in the category (that is, without having to provide them as data in the form of equations in $\mathcal{E}$).
For explicit substitution, we note that the linearity is not desirable; this is why the copy-delete comonoid $(\comonoid, \counit)$ is present: by Fox's theorem~\cite{foxCoalgebrasCartesianCategories1976}, a cartesian category is precisely a monoidal one admitting such a \emph{natural} copy-delete structure.
Thus, we must endow $\mathcal{E}$ with equations stating that $(\comonoid, \counit)$ forms a comonoid, and moreover that it is natural: $\comonoid$ behaves uniformly as a copy map, and $\counit$ behaves as deletion.

@@

As a second example, we may wish to incorporate the optimisation ${\text{plus} (x, x) \leadsto \text{mult} (x, 2)}$ in our toy theory.
This corresponds to endowing $\mathcal{E}$ with the equation of string diagrams:
\[
	\tikzfig{figures/plus-copy}
	\quad
	=
	\quad
	\tikzfig{figures/mult-2}
	\;
	.
\]
Now, in our semantic model, this rewrite can be non-destructively applied to model one step of e-graph saturation:
\[
	\tikzfig{figures/plus-copy-rewrite-1}
	\quad
	\equiv
	\quad
	\tikzfig{figures/plus-copy-rewrite-2}
	\quad
	\to
	\quad
	\tikzfig{figures/plus-copy-rewrite-3}
	\;
	.
\]
Here, $\equiv$ is a definitional equality arising from the idempotence in the Hom semilattice, and $\to$ is an application of DPO rewriting using the previous equation along the right branch.
In isolation, this isn't particularly interesting, but the important observation is that this rewriting all happens \emph{locally} in the context of a bigger diagram.

>>>>

Note that, in contrast to conventional presentations of the $\lambda$-calculus as a language of (slotted~\cite[Example 1]{slotted-egraphs}) e-graphs~\cite[Figure 10]{EggPaper}, here there is no encoding of the term formers for application, abstraction, and explicit substitution.
Instead, these constructs are handled ambiently by our semantic setting: any closed symmetric monoidal category (and in particular, $\catname{CS}(\Sigma, \mathcal{E})$ for any set of equations $\mathcal{E}$) admits an internal language in the multiplicative fragment of intuitionistic linear logic (a particular type of linear $\lambda$-calculus)~\cite[§ 1.7.3]{abramskyIntroductionCategoriesCategorical2010}.
In other words, with respect to~\autoref{fig:egraph-strings}, this allows us to \emph{natively} represent abstraction by the rounded boxes of ($\Lambda(f)$), and application by the $\textsf{ev}$ operation with $\beta$- and $\eta$-conversion arising via equality of morphisms in the category (that is, without having to provide them as data in the form of equations in $\mathcal{E}$).
For explicit substitution, we note that the linearity is not desirable; this is why the copy-delete comonoid $(\comonoid, \counit)$ is present: by Fox's theorem~\cite{foxCoalgebrasCartesianCategories1976}, a Cartesian category is precisely a monoidal one admitting such a \emph{natural} copy-delete structure.
Thus, we must endow $\mathcal{E}$ with equations stating that $(\comonoid, \counit)$ forms a comonoid, and moreover that it is natural: $\comonoid$ behaves uniformly as a copy map, and $\counit$ behaves as deletion.

@@

As a second example, we may wish to incorporate the optimisation ${\text{plus} (x, x) \leadsto \text{mult} (x, 2)}$ in our toy theory.
This corresponds to endowing $\mathcal{E}$ with the equation of string diagrams:
\[
	\tikzfig{figures/plus-copy}
	\quad
	=
	\quad
	\tikzfig{figures/mult-2}
	\;
	.
\]
Now, in our semantic model, this rewrite can be non-destructively applied to model one step of e-graph saturation:
\[\adjustbox{scale=0.8}{
	\tikzfig{figures/plus-copy-rewrite-1-2}
}
	\quad
	\equiv
	\quad
	\adjustbox{scale=0.8}{
	\tikzfig{figures/plus-copy-rewrite-2-2}
	}
	\quad
	\to
	\quad
	\adjustbox{scale=0.8}{
	\tikzfig{figures/plus-copy-rewrite-3-2}
	}
	\;
	.
\]
Here, $\equiv$ is a definitional equality arising from the idempotence in the Hom semilattice, and $\to$ is an application of DPO rewriting using the previous equation along the right branch.
In isolation, this isn't particularly interesting, but the important observation is that this rewriting all happens \emph{locally} in the context of a bigger diagram.

### 2025-07-11 POPL2026/paper/sections/soundness.tex
cosmetic changes

@@

\begin{definition}
	For closed-$\Sigma$-terms $f,g$, and a closed rewrite rule $\langle l, r \rangle$, $f \leadsto_{\langle l, r \rangle} g$ modulo the laws of symmetric monoidal categories if they are representable as either in~\autoref{def:rewrite}, or as
	$
		f = c_{1};(id_{k} \otimes d);c_{2} $ and $ g = c_{1};(id_{k} \otimes e);c_{2}
	$
	and $d \leadsto_{\langle l, r \rangle} e$
\end{definition}
\begin{lemma}
	\label{lemma:normal_form}
	Every closed $\Sigma^{+}$ term $f$ can be equivalently represented as
	$
		f_{1} + \ldots + f_{n}
	$ modulo the laws of symmetric monoidal $\catname{SLat}$-categories and distributivity law (\autoref{law:distributivity}) where none of $f_{i}$ contain the join operator.
\end{lemma}
% \begin{proof}
% Such forms are given by orienting $\catname{SLat}$-equations such that $+$ gets propagated to the top, as, for example $f \otimes (g + h) = f \otimes g + g \otimes h$.
% Most notably, we can also escape lambda abstractions by using the distributivity law (\autoref{law:distributivity}).
% \end{proof}
This takes us to the following definition of rewriting for closed $\Sigma^{+}$-terms.
\begin{definition}
	For closed $\Sigma^{+}$-term $f,g$, and a closed $\Sigma$-term rewrite rule $\langle l, r \rangle$ $f \leadsto_{\langle l, r \rangle}$ modulo the laws of symmetric monoidal $\catname{SLat}$-categories and distributivity law (\autoref{law:distributivity}) if they are representable as
	$
		f = f_{1} + \ldots + f_{i} + \ldots + f_{n} $ and $ g = f_{1} + \ldots + f_{j} + \ldots + f_{n}
	$
	such that there is a permutation $\sigma$ such that there exist indices $i,j$ and $\sigma(f_{i}) \leadsto f_{j}$.
\end{definition}
Note that the rewrite rule consist of closed $\Sigma$-terms and not closed $\Sigma^{+}$-terms, i.e. neither $l$ nor $r$ contain any $+$.
This is intentional as there rewrite rules are supposed to be domain-specific (like a rule $\langle 2 \otimes id;times, id \otimes 1; <\;\;< \rangle$) and not related to the free enrichment.

@@

\begin{proposition}[Proposition 26 and Proposition 27~\cite{fscd}]
	\label{prop:fscd}
	For two closed $\Sigma$-terms $f$ and $g$ and an $\mathcal{E}$ equation $l = r$,  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$, where the latter is defined in a sub $\catname{SLat}$-category of $\WellTypedMdaEcospans$ where carriers only contain edges from $\colorbox{yellow}{E} \cup \textcolor{closed-color}{E}$.
\end{proposition}

\begin{proposition}%
	\label{prop:quotient-structural}
	For two closed $\Sigma^{+}$-terms $f$ and $g$ and an $\mathcal{E}$ equation $l = r$,  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$.
\end{proposition}
\begin{proof}
	By definition, we have
	$
		f = f_{1} + \ldots + f_{n} $ and $  g = f_{1} + \ldots + f_{n}$,
	such that $f_{i} \leadsto f_{j}$ for some $i, j$.
	This yields two cospans of the form $\llbracket f_{1} \rrbracket + \ldots + \llbracket f_{n} \rrbracket$ for $\llbracket f \rrbracket$ and $\llbracket g \rrbracket$ and, by noting that every $f_{i}$ is a $\Sigma$-term, and carriers in every $\llbracket f_{i} \rrbracket$ have no edges from $\textcolor{e-color}{E}$, we can apply~\autoref{prop:fscd}.
\end{proof}

@@

\begin{figure}
	\[
		\adjustbox{scale=0.5}{
			\tikzfig{./figures/f_plus_g_new_2}
		}
	\]
	\captionsetup{belowskip=-1ex}
	\caption{$+$ of two morphisms in $\WellTypedMdaEcospans$}
	\label{fig:f+g}
\end{figure}

>>>>

In practice such rewrite rules are generated by the equations of the corresponding theory.

@@

\begin{definition}
	For closed-$\Sigma$-terms $f,g$, and a closed rewrite rule $\langle l, r \rangle$, $f \leadsto_{\langle l, r \rangle} g$ modulo the laws of symmetric monoidal categories if they are representable as either in~\autoref{def:rewrite}, or as
	$
		f = c_{1};(id_{k} \otimes d);c_{2} $ and $ g = c_{1};(id_{k} \otimes e);c_{2}
	$
	and $d \leadsto_{\langle l, r \rangle} e$
\end{definition}
\begin{lemma}
	\label{lemma:normal_form}
	Every closed $\Sigma^{+}$ term $f$ can be equivalently represented as
	$
		f_{1} + \ldots + f_{n}
	$ modulo the laws of symmetric monoidal $\catname{SLat}$-categories and distributivity law (\autoref{law:distributivity}) where none of $f_{i}$ contain the join operator.
\end{lemma}
% \begin{proof}
% Such forms are given by orienting $\catname{SLat}$-equations such that $+$ gets propagated to the top, as, for example $f \otimes (g + h) = f \otimes g + g \otimes h$.
% Most notably, we can also escape lambda abstractions by using the distributivity law (\autoref{law:distributivity}).
% \end{proof}
This takes us to the following definition of rewriting for closed $\Sigma^{+}$-terms.
\begin{definition}
	For closed $\Sigma^{+}$-term $f,g$, and a closed $\Sigma$-term rewrite rule $\langle l, r \rangle$ $f \leadsto_{\langle l, r \rangle}$ modulo the laws of symmetric monoidal $\catname{SLat}$-categories and distributivity law (\autoref{law:distributivity}) if they are representable as
	$
		f = f_{1} + \ldots + f_{i} + \ldots + f_{n} $ and $ g = f_{1} + \ldots + f_{j} + \ldots + f_{n}
	$
	such that there is a permutation $\sigma$ such that there exist indices $i,j$ and $\sigma(f_{i}) \leadsto f_{j}$.
\end{definition}
Note that the rewrite rule consist of closed $\Sigma$-terms and not closed $\Sigma^{+}$-terms, i.e. neither $l$ nor $r$ contain any $+$.
This is intentional as there rewrite rules are supposed to be domain-specific (like a rule $\langle 2 \otimes id;*, id \otimes 1; <\!\!< \rangle$) and not related to the free enrichment.

@@

\begin{proposition}[Proposition 26 and Proposition 27~\cite{fscd}]
	\label{prop:fscd}
	For two closed $\Sigma$-terms $f$ and $g$ and a closed rewrite rule $\langle l, r \rangle$,  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$, where the latter is defined in a sub-category of $\WellTypedMdaEcospans$ where carriers only contain edges from $\colorbox{yellow}{E} \cup \textcolor{closed-color}{E}$.
\end{proposition}

\begin{proposition}%
	\label{prop:quotient-structural}
	For two closed $\Sigma^{+}$-terms $f$ and $g$ and a closed rewrite rule $\langle l, r \rangle$,  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$.
\end{proposition}
\begin{proof}
	By definition, we have
	$
		f = f_{1} + \ldots + f_{n} $ and $  g = f_{1} + \ldots + f_{n}$,
	such that $f_{i} \leadsto f_{j}$ for some $i, j$.
	This yields two cospans of the form $\llbracket f_{1} \rrbracket + \ldots + \llbracket f_{n} \rrbracket$ for $\llbracket f \rrbracket$ and $\llbracket g \rrbracket$ and, by noting that every $f_{i}$ is a $\Sigma$-term, and carriers in every $\llbracket f_{i} \rrbracket$ have no edges from $\textcolor{e-color}{E}$, we can apply~\autoref{prop:fscd}.
\end{proof}

@@

The proposition above allows us to reason about theories of closed $\Sigma^{+}$-terms $\catname{CS}(\Sigma,\mathcal{E})^{+}$.
Given a set of equations $\mathcal{E}$ of closed $\Sigma$-terms, we construct $\llbracket \mathcal{E} \rrbracket = \{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle, \langle \llbracket r \rrbracket, \llbracket l \rrbracket \rangle \;|\; l = r \;\in\; \mathcal{E}\}$ and then we can perform DPO rewriting in $\WellTypedMdaEcospans / \mathcal{S}$ which formalises, \textit{e.g.}, the example in~\autoref{fig:e-string-substitution}.
\begin{figure}
	\[
		\adjustbox{scale=0.5}{
			\tikzfig{./figures/f_plus_g_new_2}
		}
	\]
	\captionsetup{belowskip=-1ex}
	\caption{$+$ of two morphisms in $\WellTypedMdaEcospans$}
	\label{fig:f+g}
\end{figure}
